Hippocampal replay of experience at real world speeds

### 2021-06-02 Article.tex

% Figure 1
\begin{figure}
\includegraphics[width=0.75\linewidth]{figures/Figure1/Figure1_v7}
\caption{The model can capture different sequence dynamics on simulated data. \textbf{(A)} We construct a firing sequence of 19 simulated place cells that exhibits three different movement dynamics. For the first 60 ms, one cell fires repeatedly, representing one stationary location. For next 190 ms, the cells fire in sequence, representing a rapid continuous trajectory along the virtual track. Finally, for the last 30 ms, cells fire randomly, out of spatial order, representing an fragmented spatial sequence. \textbf{(B)} Like the standard decoder, the state space model uses estimates of cells' place fields from when the animal is moving and combines them with the observed spikes in (A) to compute the likelihood of position for each time step. \textbf{(C)} The prediction from the neural data is then combined with an explicit model of each movement dynamic, which determines how latent position can change based on the position in the previous time step. We show the probability of the next position bin for each movement dynamic model (color scale). Zero here represents the previous position. \textbf{(D)} The probability of remaining in a particular movement dynamic versus switching to another dynamic is modeled as having a high probability of remaining in a particular dynamic with a small probability of switching to one of the other dynamics at each time step. \textbf{(E)} The model uses the components in A-D over all time to decode the joint posterior probability of latent position and dynamic. This can be summarized by marginalizing over latent position (left panel) to get the probability of each dynamic over time. The shaded colors indicate the category of the speed of the trajectory at that time (Stat. = Stationary, S-C-M = Stationary-Continuous-Mixture, Cont. = Continuous, F-C-M = Fragmented-Continuous-Mixture, Frag. = Fragmented), which is determined from the probability. Marginalizing the posterior across dynamics also provides an estimate of latent position over time (right panel). Red dotted line in the right panel is the best fit line from the standard decoder using the Radon transform. \textbf{(F)} The probability of each dynamic depends heavily on the speed of the trajectory, as we show using a range of simulated spiking sequences each moving at a constant speed. Each dot corresponds to the average probability of that dynamic for a given constant speed sequence. We use a 0.8 threshold (dotted line) to classify each sequence based on the dynamic or dynamics which contribute maximally to the posterior (shaded colors).
}
\label{fig:1}
% Figure 1-Supplemental 1
\figsupp[The model is robust to change of the probability of persisting in the same dynamic for a wide range of plausible expected durations (25 ms to 150 ms)]{The model is robust to change of the probability of persisting in the same dynamic for a wide range of plausible expected durations (25 ms to 150 ms). \textbf{(A)} Each plot shows the probability of each dynamic on simulated data example from \FIG{1} with a different diagonal value---which governs the probability of remaining in that dynamic. The corresponding expected duration of staying in the dynamic is listed as duration. The off-diagonal values---the probability of switching to one of the other dynamics---are set to be equally likely with the remainder of the probability, as in \FIG{1}D. The diagonal increases from left to right, top to bottom, until the case where the diagonal is one and the off-diagonal is zero---i.e. the case where there is no probability of switching to another dynamic. Shaded regions correspond to the classification as in \FIG{1}. \textbf{(B)} The probability of position over time for each diagonal value. Conventions the same as in (A).}{\includegraphics[width=1.0\linewidth]{figures/Figure1-supplemental1/Figure1_v3_supplemental1}}\label{figsupp:fig1sup1}
\end{figure}

@@

Once we have specified the data and the movement dynamic models, we have fully specified the state space model. We use acausal decoding, meaning that we use all information from both past and future spikes, to estimate the joint posterior probability of position and dynamic (see Methods). With this, we can summarize the resulting posterior probability with two quantities: the probability of each movement dynamic over time (by integrating out position; \FIG{1}E, left plot) and the probability of latent position over time, irrespective of movement dynamic (by summing over the movement dynamics; \FIG{1}E, right plot).

@@

% Figure 2-Supplemental 1
\figsupp[Decoding the same SWR in \FIG{2} with 2D position using sorted spikes and clusterless decoding.]{Decoding the same SWR in \FIG{2} with 2D position using sorted spikes and clusterless decoding. \textbf{(A)} The left plot shows the spikes from cells arranged by the linear position of the peak of place field as in \FIG{2}. The middle plot shows the probability of each dynamic over time from the 2D decode. Shaded regions indicate classification category as in \FIG{1}F and \FIG{2}. The rightmost plot shows the most probable estimate of the latent position (MAP estimate) with color indicating time. The latent position posterior summed over time is shown in the purple shading. The light grey lines represent the position of the animal over the entire recording session and the magenta dot represents the animal's position. \textbf{(B)} Same as in (A), but with clusterless decoding.}{\includegraphics[width=1.0\linewidth]{figures/Figure2-supplemental1/Figure2_v3-supplemental1}}\label{figsupp:fig2sup1}

@@

One of our criteria for a more optimal method is for it to use all of the available spiking data. Using only clustered spikes discards any spike events that cannot be uniquely assigned to a putative single neuron, substantially reducing the amount of information that the resultant decoding can use. Additionally, spike sorting is not necessary to recover the underlying neural population dynamics \citep{TrautmannAccurateEstimationNeural2019} and often requires inclusion decisions that can vary widely across experimenters. We therefore adopted a "clusterless" approach which directly relates spikes and their multiunit spike waveform features to decode position without spike sorting (see Methods). Clusterless decoding has previously been used to successfully identify theta sequences and replay sequences in the hippocampus \citep{KloostermanBayesiandecodingusing2014, ChenTransductiveneuraldecoding2012,DengRapidclassificationhippocampal2016, KayConstantSubsecondCycling2020}. Applying a clusterless decoder to the same SWR event, we get similar classification of the sequence (\FIG{2}B, D), both with 1D linearized position and 2D position (\FIGSUPP[2]{fig2sup1}B). As predicted, the spatial extent of the event is longer and the estimate of the posterior probability of latent position is narrower for the clusterless model. This reflects the clusterless model's access to a larger pool of data that provides more information about the extent of the event and more certainty in the latent position and the dynamic (\FIG{2}D vs C).

\subsection*{Most replays contain coherent spatial content}
After testing our model on a single SWR event, we applied our clusterless decoding algorithm to hippocampal recordings from 10 rats performing the W-track spatial alternation task (\# tetrodes range: [10, 24], brain areas = [CA1, CA2, CA3]; some data previously used in \cite{KarlssonAwakereplayremote2009, KayConstantSubsecondCycling2020, CarrTransientSlowGamma2012}, position linearized to 1D). We first confirmed that our cell population for each animal could closely track the position of the animal during behavior by comparing the most probable decoded position to the position of the animal. Even including theta sequences and replay sequences, which can represent positions other than where the animal is, we found that the average median difference between actual and decoded location is small (4 cm median difference, 3-6 cm 95\% CI, for all time in sessions, 5 fold cross validation). This level of accuracy is just above the size of our 3 cm position bins, and is comparable to that of other studies \citep{DavidsonHippocampalReplayExtended2009, ShinDynamicsAwakeHippocampalPrefrontal2019, FarooqEmergencepreconfiguredplastic2019}. Importantly, we achieved this using small time bins (2 ms) compared to the commonly used 250 ms time bin that most studies use for decoding during movement, demonstrating the power of our algorithm. For each animal, we also verified that the population spiking rate was relatively consistent over all positions on the track, suggesting that we have comparable sampling of all locations in the environment (\FIGSUPP[2]{fig2sup3}).

@@

% Figure 3-Supplemental 1
\figsupp[More examples of SWRs with non-constant speed trajectories]{A-I. More examples of SWRs that would not be well-characterized by using the standard decoder. Conventions are the same as in \FIG{3}.}{\includegraphics[width=1.0\linewidth]{figures/Figure3-supplemental1/Figure3_v4_supplemental1}}\label{figsupp:fig3sup1}

@@

Because our model is specified in the context of a latent position associated with different movement dynamics, it allows us to not only classify events in terms of their dynamics, but also to quantify the model's certainty in each position estimate at each moment in time given the model parameters. To do so, we can compute the cumulative spatial size of the 95\% highest posterior density (HPD) region of the latent position estimate. The 95\% HPD region corresponds to the set of most probable latent positions as estimated by the model. Larger values of the HPD region size indicate the model is less certain about position, because the most probable decoded positions are distributed over more of the track at a given time point. In contrast, smaller values indicate that the model is more certain about the estimate of position because the extent of the HPD is more concentrated and covers less of the track. Thus, the HPD region provides a complementary measure of spatial information, and evaluating it allows us to verify that the events we defined as spatially coherent also correspond to events where there is high certainty around the position estimates.

@@

% Figure 4-Supplemental 1
\figsupp[Average 95\% HPD size increases on the 95\% highest posterior density on position shuffled data]{Comparison of average spatial coverage of the 95\% highest posterior density on real vs. position shuffled data for two recording sessions from different animals. Red line represents the average spatial position spanned by the highest posterior density on real data for all ripples. The histogram represents the distribution after 50 shuffles of the position data. Position data was shuffled by resampling with replacement from the set of all observed positions in that recording session, destroying position information but preserving spiking timing and position occupancy.}{\includegraphics[width=0.8\linewidth]{figures/Figure4-supplemental1/Figure4_v1-supplemental1}}\label{figsupp:fig4sup1}

% Figure 4-Supplemental 2
\figsupp[Average 95\% HPD size increases on run-swapped, circularly shuffled position data]{Comparison of average spatial coverage of the 95\% highest posterior density on real vs. run-swapped, circularly shuffled position data for two recording sessions from different animals. Red line represents the average spatial position spanned by the highest posterior density on real data for all ripples. The histogram represents the distribution after 50 shuffles of the position data. The histogram represents the distribution after 50 shuffles of the run from well to well as well as the circularly shuffled position (for each tetrode). This preserves local spatial correlations in the data but breaks the global spatial relationships between the spikes and the data.}{\includegraphics[width=0.8\linewidth]{figures/Figure4-supplemental2/Figure4_v1-supplemental2}}\label{figsupp:fig4sup2}

@@

Finally, we asked how our method compares to the standard approaches to identifying significant replay events via a non-parametric shuffle test and characterizing their content. This comparison is complicated due to the variety of parameters and inference methods used previously, so here we chose three commonly used approaches and a reasonable parameter set for the comparison. We began with an examination of events based on statistical significance of a line fit to a circularly permuted spatial distribution, which has been the standard in the field.  We applied the clusterless algorithm in 20 ms bins to produce probability densities of represented locations, as this binning is comparable to previous methods. We then used the radon transform method, which finds the line with the highest summed posterior probability of position along the line \citep{DavidsonHippocampalReplayExtended2009, FarooqEmergencepreconfiguredplastic2019}, to calculate, for each SWR, a measure of significance based on the assumption that the probability density moves at a constant speed throughout the event. We found that \texttildelow12\% of the classified events (as defined by our state-space decoder) were identified as significant. Not surprisingly, the fraction of significant events was strongly related to which dynamic they contained. Of the SWRs that contained a period of continuous dynamics, 31\% (1386 of 4511 SWRs) were significant. Of all SWRs that contained a period of stationary dynamics, 8\% (542 of 7176 SWRs) were significant, and when we restricted the analysis to long stationary events (>100 ms duration), we found that 28\% were significant. This increase is not surprising, since the shuffle used for the radon method penalizes shorter events. An alternative approach, the linear regression method, which samples from the posterior probability and uses linear regression to fit a line to the samples \citep{KarlssonAwakereplayremote2009, CarrHippocampalreplayawake2011, ShinDynamicsAwakeHippocampalPrefrontal2019}, detects only non-zero slopes as significant, and thus by definition cannot identify stationary events as significant.

@@

Previous work reported that less than half of SWRs or high-population-activity events contained replay events \citep{KarlssonAwakereplayremote2009, FosterReversereplaybehavioural2006, DavidsonHippocampalReplayExtended2009}. This could lead one to the assumption that most events are not spatially coherent or meaningful. Our results indicate that this is not the case: in our dataset, the large majority of SWRs (86\%) contained spatially coherent content, as defined by some combination of stationary and continuous dynamics. These events were mostly decoded representations where the most probable positions were highly spatially concentrated (i.e. had small HPD size), indicating that they were driven by spatially specific firing. This was in contrast to unclassified events or events that included fragmented dynamics, which were associated with highly distributed decoded spatial representations.

@@

The different questions addressed by our method compared to previous methods can also explain the apparent differences in findings. Our results showed that 19\% of SWRs contained continuous trajectories---meaning that portions of the SWR had spatially coherent trajectories that moved at high velocities---but only 4\% were continuous throughout the entire SWR. This, \textit{prima facie}, may seem at odds with previous results, which found between 5\% and 45\% of SWRs were "significant", meaning that they were well fit by a line compared to various shuffled versions of the dataset. Furthermore, these studies found trajectories on average move at much higher speeds (20x normal rat movement speed) than we report here. These previous findings were based largely on a single linear fits to the data for each event, while our model allows the trajectory to change speed during the SWR, which can lead to very different estimates of the speed of the event. Many of the events that had continuous dynamics, and would be classified as having high speed trajectories, also exhibited stationary-continuous mixture dynamics for some portion of the event(see for example \FIG{1}). Further, because the linear fit includes the whole SWR event, it is highly sensitive to the definition of the start and end of the SWR event, an often arbitrary definition which varies across research groups. In contrast, our model makes an estimate for each time point and only depends on the previous and future time step, which minimizes the dependence of the event on those boundaries and any possible "noise" activity within the SWR. This, along with our more permissive threshold for SWRs, allows us to characterize more SWRs than previous studies, rather than exclude them from analysis.

@@

State-space models like the one we use here can enable a richer set of analyses of events that take advantage of all of the available data. These approaches can be applied not just to SWRs but to all times, providing a moment-by-moment estimate of the nature of the spatial representation in the hippocampus, including important information about the spatial coherence of that representation. The model can be extended to incorporate other previously experienced environments by training place field models on those environments and including the appropriate movement transition matrices for those environments. It can also be extended to account for task and forward/reverse sequences as in \cite{DengRapidclassificationhippocampal2016}. Finally, models can be built to estimate any covariate, including movement direction \citep{KayConstantSubsecondCycling2020}. We therefore suggest that this approach has the potential to provide a much richer understanding of neural representations throughout the brain.

@@

\subsection*{Highest Posterior Density}
The 95\% highest posterior density (HPD) is a measurement of the spread of the posterior probability and is defined as the region of the posterior that contains the top 95\% of values of the posterior probability \citep{CasellaStatisticalinference2001}. By using only the top values, this measurement of spread is not influenced by multimodal distributions (whereas an alternative measure like the quantiles of the distribution would be). In this manuscript, we use the HPD size---the total area of the track covered by the 95\% HPD region---to evaluate the uncertainty of the posterior probability of position.

In order to calculate the HPD size, we first calculate the 95\% HPD region by determining the maximum threshold value $h$ that fulfills the follow equality:
$$
\int_{\{x: p(x_{k} \mid O_{1:T}) > h\}} p(x_{k} \mid O_{1:T})dx = 0.95
$$
The 95\% HPD region is the set of position bins with posterior values greater than the threshold $\{x : p(x_{k} \mid O_{1:T}) > h\}$. The HPD size is calculated by taking the integral of the members of this set:
$$
\int_{\{x: p(x_{k} \mid O_{1:T}) > h\}} \mathbb{1}dx
$$
which we approximate with a Riemann sum.

\subsection*{Shuffle Analysis of the Effect of Place Encoding on Classification of Dynamics and HPD Size}
In order to confirm that the model classification of dynamics depended on hippocampal place specific encoding, we resampled the position during movement with replacement, but preserved the spike times and spike waveform amplitudes. We fit the clusterless encoding model on the resampled data and then decoded the immobility periods. Like with the non-resampled data, we then extracted the SWR times and determined their classification based on our classification scheme. We repeated this shuffle analysis 50 times and compared this distribution to the real data for two recording sessions on two different animals (animal Bond, day 3, epoch 2 and animal Remy, day 35, epoch 2).

@@

\subsection*{Software and Code Availability}
Python code used for analysis and generating figures in the paper is available at: \url{https://github.com/Eden-Kramer-Lab/replay_trajectory_paper}. Code for the classifier is available in a separate software repository to facilitate code reuse at: \url{https://github.com/Eden-Kramer-Lab/replay_trajectory_classification} (doi: 10.5281/zenodo.3713412).
Both code bases rely on the following python packages: Numpy \citep{vanderWaltNumPyArrayStructure2011}, Numba \citep{LamNumbaLLVMbasedPython2015}, Matplotlib \citep{HunterMatplotlib2DGraphics2007}, xarray \citep{HoyerxarrayNDlabeled2017}, NetworkX \citep{HagbergExploringNetworkStructure2008}, Pandas \citep{McKinneyDataStructuresStatistical2010} and Seaborn \citep{Waskomseabornstatisticaldata2021}. All code is open-source and licensed under the MIT Software License. Classifier code can be easily installed as a python package with all requisite dependencies using pip or conda. See software repositories for specific details.

>>>>

% Figure 1
\begin{figure}
\includegraphics[width=0.75\linewidth]{figures/Figure1/Figure1_v7}
\caption{The model can capture different sequence dynamics on simulated data. \textbf{(A)} We construct a firing sequence of 19 simulated place cells that exhibits three different movement dynamics. For the first 60 ms, one cell fires repeatedly, representing one stationary location. For next 190 ms, the cells fire in sequence, representing a rapid continuous trajectory along the virtual track. For the last 30 ms, cells fire randomly, out of spatial order, representing an fragmented spatial sequence. \textbf{(B)} Like the standard decoder, the state space model uses estimates of cells' place fields from when the animal is moving and combines them with the observed spikes in (A) to compute the likelihood of position for each time step. \textbf{(C)} The prediction from the neural data is then combined with an explicit model of each movement dynamic, which determines how latent position can change based on the position in the previous time step. We show the probability of the next position bin for each movement dynamic model (color scale). Zero here represents the previous position. \textbf{(D)} The probability of remaining in a particular movement dynamic versus switching to another dynamic is modeled as having a high probability of remaining in a particular dynamic with a small probability of switching to one of the other dynamics at each time step. \textbf{(E)} The model uses the components in A-D over all time to decode the joint posterior probability of latent position and dynamic. This can be summarized by marginalizing over latent position (left panel) to get the probability of each dynamic over time. The shaded colors indicate the category of the speed of the trajectory at that time (Stat. = Stationary, S-C-M = Stationary-Continuous-Mixture, Cont. = Continuous, F-C-M = Fragmented-Continuous-Mixture, Frag. = Fragmented), which is determined from the probability. Marginalizing the posterior across dynamics also provides an estimate of latent position over time (right panel). Red dotted line in the right panel is the best fit line from the standard decoder using the Radon transform. \textbf{(F)} The probability of each dynamic depends heavily on the speed of the trajectory, as we show using a range of simulated spiking sequences each moving at a constant speed. Each dot corresponds to the average probability of that dynamic for a given constant speed sequence. We use a 0.8 threshold (dotted line) to classify each sequence based on the dynamic or dynamics which contribute maximally to the posterior (shaded colors).
}
\label{fig:1}
% Figure 1-Supplemental 1
\figsupp[The model is robust to change of the probability of persisting in the same dynamic for a wide range of plausible expected durations (25 ms to 150 ms)]{The model is robust to change of the probability of persisting in the same dynamic for a wide range of plausible expected durations (25 ms to 150 ms). \textbf{(A)} Each panel shows the probability of each dynamic on simulated data example from \FIG{1} with a different diagonal value---which governs the probability of remaining in that dynamic. The corresponding expected duration of staying in the dynamic is listed as duration. The off-diagonal values---the probability of switching to one of the other dynamics---are set to be equally likely with the remainder of the probability, as in \FIG{1}D. The diagonal increases from left to right, top to bottom, until the case where the diagonal is one and the off-diagonal is zero---i.e. the case where there is no probability of switching to another dynamic. Shaded regions correspond to the classification as in \FIG{1}F. \textbf{(B)} The probability of position over time for each diagonal value. Conventions the same as in (A).}{\includegraphics[width=1.0\linewidth]{figures/Figure1-supplemental1/Figure1_v3_supplemental1}}\label{figsupp:fig1sup1}
\end{figure}

@@

Once we have specified the data and the movement dynamic models, we have fully specified the state space model. We use acausal decoding, meaning that we use all information from both past and future spikes, to estimate the joint posterior probability of position and dynamic (see Methods). With this, we can summarize the resulting posterior probability with two quantities: the probability of each movement dynamic over time (by integrating out position; \FIG{1}E, left panel) and the probability of latent position over time, irrespective of movement dynamic (by summing over the movement dynamics; \FIG{1}E, right panel).

@@

% Figure 2-Supplemental 1
\figsupp[Decoding the same SWR in \FIG{2} with 2D position using sorted spikes and clusterless decoding.]{Decoding the same SWR in \FIG{2} with 2D position using sorted spikes and clusterless decoding. \textbf{(A)} The left panel shows the spikes from cells arranged by the linear position of the peak of place field as in \FIG{2}. The middle panel shows the probability of each dynamic over time from the 2D decode. Shaded regions indicate classification category as in \FIG{1}F and \FIG{2}. The rightmost panel shows the most probable estimate of the latent position (MAP estimate) with color indicating time. The latent position posterior summed over time is shown in the purple shading. The light grey lines represent the position of the animal over the entire recording session and the magenta dot represents the animal's position. \textbf{(B)} Same as in (A), but with clusterless decoding.}{\includegraphics[width=1.0\linewidth]{figures/Figure2-supplemental1/Figure2_v3-supplemental1}}\label{figsupp:fig2sup1}

@@

One of our criteria for a more optimal method is for it to use all of the available spiking data. Using only clustered spikes discards any spike events that cannot be uniquely assigned to a putative single neuron, substantially reducing the amount of information that the resultant decoding can use. Additionally, spike sorting is not necessary to recover the underlying neural population dynamics \citep{TrautmannAccurateEstimationNeural2019} and often requires inclusion decisions that can vary widely across experimenters. We therefore adopted a "clusterless" approach which directly uses multiunit spikes and their spike waveform features to decode position without spike sorting (see Methods). Clusterless decoding has previously been used to successfully identify theta sequences and replay sequences in the hippocampus \citep{KloostermanBayesiandecodingusing2014, ChenTransductiveneuraldecoding2012,DengRapidclassificationhippocampal2016, KayConstantSubsecondCycling2020}. Applying a clusterless decoder to the same SWR event, we get similar classification of the sequence (\FIG{2}B, D), both with 1D linearized position and 2D position (\FIGSUPP[2]{fig2sup1}B). As predicted, the spatial extent of the event is longer and the estimate of the posterior probability of latent position is narrower for the clusterless model. This reflects the clusterless model's access to a larger pool of data that provides more information about the extent of the event and more certainty in the latent position and the dynamic (\FIG{2}D vs C).

\subsection*{Most replays contain coherent spatial content}
After testing our model on a single SWR event, we applied our clusterless decoding algorithm to hippocampal recordings from 10 rats performing the W-track spatial alternation task (\# tetrodes range: [10, 24], brain areas = [CA1, CA2, CA3]; some data previously used in \cite{KarlssonAwakereplayremote2009, KayConstantSubsecondCycling2020, CarrTransientSlowGamma2012}; position linearized to 1D). We first confirmed that our cell population for each animal could closely track the position of the animal during behavior by comparing the most probable decoded position to the position of the animal. Even including theta sequences and replay sequences, which can represent positions other than where the animal is, we found that the average median difference between actual and decoded location is small (4 cm median difference, 3-6 cm 95\% CI, for all time in sessions, 5 fold cross validation). This level of accuracy is just above the size of our 3 cm position bins, and is comparable to that of other studies \citep{DavidsonHippocampalReplayExtended2009, ShinDynamicsAwakeHippocampalPrefrontal2019, FarooqEmergencepreconfiguredplastic2019}. Importantly, we achieved this using small time bins (2 ms) compared to the commonly used 250 ms time bin that most studies use for decoding during movement, demonstrating the power of our algorithm. For each animal, we also verified that the population spiking rate was relatively consistent over all positions on the track, suggesting that we have comparable sampling of all locations in the environment (\FIGSUPP[2]{fig2sup3}).

@@

% Figure 3-Supplemental 1
\figsupp[More examples of SWRs with non-constant speed trajectories]{A-I. More examples of SWRs with non-constant speed trajectories. Conventions are the same as in \FIG{3}.}{\includegraphics[width=1.0\linewidth]{figures/Figure3-supplemental1/Figure3_v4_supplemental1}}\label{figsupp:fig3sup1}

@@

Because our model is specified in the context of a latent position associated with different movement dynamics, it allows us to not only classify events in terms of their dynamics, but also to quantify the model's certainty in each position estimate at each moment in time given the model parameters. To do so, we can compute the cumulative spatial size of the 95\% highest posterior density (HPD) region of the latent position estimate. The 95\% HPD region corresponds to the set of most probable latent positions as estimated by the model. Larger values of the HPD region size indicate the model is less certain about position, because the most probable decoded positions are distributed over more of the track at a given time point. In contrast, smaller values of the HPD region size indicate that the model is more certain about the estimate of position because the extent of the HPD is more concentrated and covers less of the track. Thus, the HPD region size provides a complementary measure of spatial information, and evaluating it allows us to verify that the events we defined as spatially coherent also correspond to events where there is high certainty around the position estimates.

@@

% Figure 4-Supplemental 1
\figsupp[Shuffling the position with replacement increases the average 95\% HPD region size]{Average 95\% HPD region size increases on position shuffled data for two recording sessions from different animals. Red line represents the average 95\% HPD region size for all ripples. The histogram represents the distribution after 50 shuffles of the position data. Position data was shuffled by resampling with replacement from the set of all observed positions in that recording session, destroying position information but preserving spiking timing and position occupancy.}{\includegraphics[width=0.8\linewidth]{figures/Figure4-supplemental1/Figure4_v1-supplemental1}}\label{figsupp:fig4sup1}

% Figure 4-Supplemental 2
\figsupp[Shuffling the position data increases the average 95\% HPD region size for spatially coherent and incoherent classified times]{Average 95\% HPD region size increases for run-swapped, circularly shuffled position data compared to real data. Data from two recording sessions from different animals. Red line represents the average spatial position spanned by the highest posterior density on real data for all ripples. The histogram represents the distribution after 50 shuffles of the position data. The histogram represents the distribution after 50 shuffles of the run from well to well as well as the circularly shuffled position (for each tetrode). This preserves local correlations between spikes but reduces global spatial structure.}{\includegraphics[width=0.8\linewidth]{figures/Figure4-supplemental2/Figure4_v1-supplemental2}}\label{figsupp:fig4sup2}

@@

Finally, we asked how our method compares to the standard approaches to identifying significant replay events via a non-parametric shuffle test and characterizing their content. This comparison is complicated due to the variety of parameters and inference methods used previously, so here we chose three commonly used approaches and a reasonable parameter set for the comparison. We began with an examination of events based on statistical significance of a line fit to a circularly permuted spatial distribution, which has been the standard in the field.  We applied the clusterless algorithm in 20 ms bins to produce probability densities of represented locations, as this binning is comparable to previous methods. We then used the radon transform method, which finds the line with the highest summed posterior probability of position along the line \citep{DavidsonHippocampalReplayExtended2009, FarooqEmergencepreconfiguredplastic2019}, to calculate, for each SWR, a measure of significance based on the assumption that the probability density moves at a constant speed throughout the event. We found that \texttildelow12\% of the classified events (as defined by our state-space decoder) were identified as significant. Not surprisingly, the fraction of significant events was strongly related to which dynamic they contained. Of the SWRs that contained a period of continuous dynamics, 31\% (1,386 of 4,511 SWRs) were significant. Of all SWRs that contained a period of stationary dynamics, 8\% (542 of 7,176 SWRs) were significant, and when we restricted the analysis to long stationary events (>100 ms duration), we found that 28\% were significant. This increase is not surprising, since the shuffle used for the radon method penalizes shorter events. An alternative approach, the linear regression method, which samples from the posterior probability and uses linear regression to fit a line to the samples \citep{KarlssonAwakereplayremote2009, CarrHippocampalreplayawake2011, ShinDynamicsAwakeHippocampalPrefrontal2019}, detects only non-zero slopes as significant, and thus by definition cannot identify stationary events as significant.

@@

Previous work reported that less than half of SWRs or high-population-activity events contained replay events \citep{KarlssonAwakereplayremote2009, FosterReversereplaybehavioural2006, DavidsonHippocampalReplayExtended2009}. This could lead one to the assumption that most events are not spatially coherent or meaningful. Our results indicate that this is not the case: in our dataset, the large majority of SWRs (86\%) contained spatially coherent content, as defined by some combination of stationary and continuous dynamics. These events were mostly decoded representations where the most probable positions were highly spatially concentrated (i.e. had small HPD region size), indicating that they were driven by spatially specific firing. This was in contrast to unclassified events or events that included fragmented dynamics, which were associated with highly distributed decoded spatial representations.

@@

The different questions addressed by our method compared to previous methods can also explain the apparent differences in findings. Our results showed that 19\% of SWRs contained continuous trajectories---meaning that portions of the SWR had spatially coherent trajectories that moved at high velocities---but only 4\% were continuous throughout the entire SWR. This, \textit{prima facie}, may seem at odds with previous results, which found between 5\% and 45\% of SWRs were "significant", meaning that they were well fit by a line compared to various shuffled versions of the dataset. Furthermore, these studies found trajectories on average move at much higher speeds (20x normal rat movement speed) than we report here. These previous findings were based largely on a single linear fits to the data for each event, while our model allows the trajectory to change speed during the SWR, which can lead to very different estimates of the speed of the event. Many of the events that had continuous dynamics, and would be classified as having high speed trajectories, also exhibited stationary-continuous mixture dynamics for some portion of the event(see for example \FIG{2}). Further, because the linear fit includes the whole SWR event, it is highly sensitive to the definition of the start and end of the SWR event, an often arbitrary definition which varies across research groups. In contrast, our model makes an estimate for each time point and only depends on the previous and future time step, which minimizes the dependence of the event on those boundaries and any possible "noise" activity within the SWR. This, along with our more permissive threshold for SWRs, allows us to characterize more SWRs than previous studies, rather than exclude them from analysis.

@@

State-space models like the one we use here can enable a richer set of analyses of events that take advantage of all of the available data. These approaches can be applied not just to SWRs but to all times, providing a moment-by-moment estimate of the nature of the spatial representation in the hippocampus, including important information about the spatial coherence of that representation. The model can be extended to incorporate other previously experienced environments by training place field models on those environments and including the appropriate movement transition matrices for those environments. It can also be extended to account for task conditions (such as the inbound and outbound conditions in our spatial alternation task) and forward/reverse sequences as in \cite{DengRapidclassificationhippocampal2016}. Finally, models can be built to estimate any covariate, including movement direction \citep{KayConstantSubsecondCycling2020}. We therefore suggest that this approach has the potential to provide a much richer understanding of neural representations throughout the brain.

@@

\subsection*{Highest Posterior Density}
The 95\% highest posterior density (HPD) is a measurement of the spread of the posterior probability and is defined as the region of the posterior that contains the top 95\% of values of the posterior probability \citep{CasellaStatisticalinference2001}. By using only the top values, this measurement of spread is not influenced by multimodal distributions (whereas an alternative measure like the quantiles of the distribution would be). In this manuscript, we use the HPD region size---the total area of the track covered by the 95\% HPD region---to evaluate the uncertainty of the posterior probability of position.

In order to calculate the HPD region size, we first calculate the 95\% HPD region by determining the maximum threshold value $h$ that fulfills the follow equality:
$$
\int_{\{x: p(x_{k} \mid O_{1:T}) > h\}} p(x_{k} \mid O_{1:T})dx = 0.95
$$
The 95\% HPD region is the set of position bins with posterior values greater than the threshold $\{x : p(x_{k} \mid O_{1:T}) > h\}$. The HPD region size is calculated by taking the integral of the members of this set:
$$
\int_{\{x: p(x_{k} \mid O_{1:T}) > h\}} \mathbb{1}dx
$$
which we approximate with a Riemann sum.

\subsection*{Shuffle Analysis of the Effect of Place Encoding on Classification of Dynamics and HPD region size}
In order to confirm that the model classification of dynamics depended on hippocampal place specific encoding, we resampled the position during movement with replacement, but preserved the spike times and spike waveform amplitudes. We fit the clusterless encoding model on the resampled data and then decoded the immobility periods. Like with the non-resampled data, we then extracted the SWR times and determined their classification based on our classification scheme. We repeated this shuffle analysis 50 times and compared this distribution to the real data for two recording sessions on two different animals (animal Bond, day 3, epoch 2 and animal Remy, day 35, epoch 2).

@@

\subsection*{Software and Code Availability}
Python code used for analysis and generating figures in the paper is available at: \url{https://github.com/Eden-Kramer-Lab/replay_trajectory_paper}. Code for the decoder is available in a separate software repository to facilitate code reuse at: \url{https://github.com/Eden-Kramer-Lab/replay_trajectory_classification} (doi: 10.5281/zenodo.3713412).
Both code bases rely on the following python packages: Numpy \citep{vanderWaltNumPyArrayStructure2011}, Numba \citep{LamNumbaLLVMbasedPython2015}, Matplotlib \citep{HunterMatplotlib2DGraphics2007}, xarray \citep{HoyerxarrayNDlabeled2017}, NetworkX \citep{HagbergExploringNetworkStructure2008}, Pandas \citep{McKinneyDataStructuresStatistical2010} and Seaborn \citep{Waskomseabornstatisticaldata2021}. All code is open-source and licensed under the MIT Software License. Decoder code can be easily installed as a python package with all requisite dependencies using pip or conda. See software repositories for specific details.
